---
title: "Trapezoid Rule and Monte Carlo Estimation"
author: "Michael Rose"
date: "October 29, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tibble)
library(ggforce)
library(dplyr)
library(magrittr)
options(warn=-1)
```

# Trapezoidal Rule with 10 Partitions

Consider $\int_{0}^1 x^2 dx$. Using evaluation at 10 points, we will approximate it using the trapezoid rule in **R**. 

The composite trapezoidal rule divides the integral into $n$ subintervals. The trapezoid rule is then performed on each of those $n$ subintervals. 

Let our function $f$ be twice differentiable in the interval $[a, b]$. Also let $h = \frac{(b - a)}{n}$ and $x_j = a + jh$ for each $j = 0, 1, ..., n$. Then the composite trapezoidal rule, with its error term is defined as the following: 

$\int_a^b f(x) dx = \frac{h}{2}[f(a) + 2 \sum_{j=1}^{n-1} f(x_j) + f(b)] - \frac{b-a}{12} h^2 f''(\mu)$

where, since there exists a number $\mu$ between $[a, b]$:

$\mathrm{error} = - \frac{b-a}{12n^2} f''(\mu)$. 

```{r}
# define our function 
given_function <- function(x){
  return(x * x)
}

# break the interval into 10 subintervals 
subintervals <- seq.int(0, 1, length.out = 10)

# create a vector for functions outputs 
fx <- vector(length = length(subintervals))

# for each subinterval, calculate the function 
for (i in 1:length(subintervals)){
  fx[i] <- given_function(subintervals[i])
}

# collect our points into a data frame 
needed_points <- tibble(xend = subintervals, 
                     y = rep(0, 10), 
                     yend = fx,
                     yend1 = c(fx[2:10], fx[10]), 
                     xend1 = c(subintervals[2:10], subintervals[10])
                     )

# plot the function and its approximation 
ggplot(data = needed_points) + 
  stat_function(fun = given_function, size = 1.05, alpha = 0.75, color = 'lightskyblue3') + 
  geom_segment(aes(x = xend, y = y, xend = xend, yend = yend)) + 
  geom_segment(aes(x = xend, y= yend, xend = xend1, yend = yend1)) + 
  geom_ribbon(aes(x = xend, ymin = y, ymax = yend), fill = 'royalblue4', alpha = 0.8) + 
  geom_area(stat = 'function', fun = given_function, fill = 'black', alpha = 0.3, xlim = c(0, 1)) + 
  xlim(c(0, 1.25)) + ylim(c(0, 1.25))

# zooming in
ggplot(data = needed_points) + 
  stat_function(fun = given_function, size = 1.05, alpha = 0.75, color = 'lightskyblue3') + 
  geom_segment(aes(x = xend, y = y, xend = xend, yend = yend)) + 
  geom_segment(aes(x = xend, y= yend, xend = xend1, yend = yend1)) + 
  geom_ribbon(aes(x = xend, ymin = y, ymax = yend), fill = 'royalblue4', alpha = 0.8) + 
  geom_area(stat = 'function', fun = given_function, fill = 'black', alpha = 0.3, xlim = c(0, 1)) + 
  xlim(c(0, 0.35)) + ylim(c(0, 0.25))
```

# Numerical Solution 

```{r}
# composite trapezoid function 
comp_trapezoid <- function(f, a, b, n){
  # check to make sure the function f is valid 
  if (is.function(f) == FALSE){
    stop('f must be a valid function with one parameter.')
  }
  # implementation
  h <- (b-a)/n
  j <- 1:n - 1
  xj <- a + j*h
  approx <- (h/2) * (f(a) + 2 * sum(f(xj)) + f(b))
  
  return(approx)
}

# calculation with 10 partitions
comp_trapezoid(given_function, 0, 1, 10)

```

# Trapezoidal Rule with 100 Partitions 

```{r}
# break the interval into 100 subintervals 
subintervals <- seq.int(0, 1, length.out = 100)

# create a vector for functions outputs 
fx <- vector(length = length(subintervals))

# for each subinterval, calculate the function 
for (i in 1:length(subintervals)){
  fx[i] <- given_function(subintervals[i])
}

# collect our points into a data frame 
needed_points <- tibble(xend = subintervals, 
                     y = rep(0, 100), 
                     yend = fx,
                     yend1 = c(fx[2:100], fx[100]), 
                     xend1 = c(subintervals[2:100], subintervals[100])
                     )

# plot the function and its approximation 
ggplot(data = needed_points) + 
  stat_function(fun = given_function, size = 1.05, alpha = 0.75, color = 'lightskyblue3') + 
  geom_segment(aes(x = xend, y = y, xend = xend, yend = yend)) + 
  geom_segment(aes(x = xend, y= yend, xend = xend1, yend = yend1)) + 
  geom_ribbon(aes(x = xend, ymin = y, ymax = yend), fill = 'royalblue4', alpha = 0.8) + 
  geom_area(stat = 'function', fun = given_function, fill = 'black', alpha = 0.3, xlim = c(0, 1)) + 
  xlim(c(0, 1.25)) + ylim(c(0, 1.25))

# zooming in
ggplot(data = needed_points) + 
  stat_function(fun = given_function, size = 1.05, alpha = 0.75, color = 'lightskyblue3') + 
  geom_segment(aes(x = xend, y = y, xend = xend, yend = yend)) + 
  geom_segment(aes(x = xend, y= yend, xend = xend1, yend = yend1)) + 
  geom_ribbon(aes(x = xend, ymin = y, ymax = yend), fill = 'royalblue4', alpha = 0.8) + 
  geom_area(stat = 'function', fun = given_function, fill = 'black', alpha = 0.3, xlim = c(0, 1)) + 
  xlim(c(0, 0.35)) + ylim(c(0, 0.25))

# closer 
ggplot(data = needed_points) + 
  stat_function(fun = given_function, size = 1.05, alpha = 0.75, color = 'lightskyblue3') + 
  geom_segment(aes(x = xend, y = y, xend = xend, yend = yend)) + 
  geom_segment(aes(x = xend, y= yend, xend = xend1, yend = yend1)) + 
  geom_ribbon(aes(x = xend, ymin = y, ymax = yend), fill = 'royalblue4', alpha = 0.8) + 
  geom_area(stat = 'function', fun = given_function, fill = 'black', alpha = 0.3, xlim = c(0, 1)) + 
  xlim(c(0, 0.15)) + ylim(c(0, 0.15))
```


```{r}
# calculation with 100 partitions 
comp_trapezoid(given_function, 0, 1, 100)
```


# Monte Carlo Approximation 

We wish to integrate $I(f) = \int_a^b f(x) dx$. 

* First we will choose some pdf we hope to sample from $g(x) \in [a, b]$. 
* Then we can generate data $X_1, X_2, ..., X_n$ from $g(x)$. 
* Finally, we estimate $I(f) = \sum_{i = 1}^{n} \frac{f(g(x))}{n}$ and each iteration provides a new sample from $g(x)$. 

```{r}
# function for arbitrary f
# n is number iterations
# a is lower bound 
# b is upper bound
# f is the function to be approximated
# since f in [a, b] we can use a uniform sampling function

mc_Integral_Unif <- function(n, a, b, f){
  # check to make sure functions are valid
  if (is.function(f) == FALSE){
    stop('f and g must be valid functions with one parameter.')
  }
  
  # sampling function
  g <- runif(n, a, b)
  
  # 
  y <- (f(g))/(1/(b-a))
  
  # int approximation 
  Int <- sum(y)/n
  
  # standard error of int and 95% confidence interval 
  y_squared <- y^2
  se <- sqrt((sum(y_squared)/n-Int^2)/n)
  ci_l <- Int - 1.96 * se
  ci_u <- Int + 1.96 * se
  list("Int" = Int, "SE" = se, "95% CI Lower Bound" = ci_l, "95% CI Upper Bound" = ci_u)
}
```

```{r}
# run for 10 iterations
mc_Integral_Unif(10, 0, 1, given_function)

# run for 100 iterations
mc_Integral_Unif(100, 0, 1, given_function)
```

### Higher Iterations of Monte Carlo

```{r}

# break the interval into 100 subintervals 
subintervals <- seq.int(0, 1000000, length.out = 100)

# create a matrix for functions outputs 
fx <- matrix(NA, nrow = length(subintervals), ncol = 4)

# for each subinterval, calculate the function 
for (i in 1:length(subintervals)){
  v <- mc_Integral_Unif(subintervals[i], 0, 1, given_function)
  fx[i, 1] <- v$Int
  fx[i, 2] <- v$SE
  fx[i, 3] <- v$`95% CI Lower Bound`
  fx[i, 4] <- v$`95% CI Upper Bound`
}

# collect our points into a data frame 
needed_points <- tibble(
                     exp_num = 1:100, 
                     num_iters = subintervals,
                     approx_val = fx[, 1],
                     standard_error = fx[, 2],
                     conf_int_lb = fx[, 3], 
                     conf_int_ub = fx[,4]
                     )

needed_points
```

```{r}
# value vs num iters
ggplot(needed_points, aes(needed_points$num_iters, needed_points$approx_val)) +
  geom_point() +
  geom_hline(yintercept = 1/3, color = "royalblue4") +
  ylab("Approximated Value") + xlab("Number of Iterations of MC") +
  ylim(c(0.32, 0.34))

# value vs standard error 
ggplot(needed_points, aes(needed_points$standard_error, needed_points$approx_val)) +
  geom_point() +
  geom_hline(yintercept = 1/3, color = "royalblue4") +
  ylab("Approximated Value") + xlab("Standard Error") +
  ylim(c(0.32, 0.34)) + scale_x_reverse()

# look at points with less than 5/10,000th error
se_0005 <- needed_points %>% 
  filter(needed_points$standard_error < 0.0005)

ggplot(se_0005, aes(se_0005$standard_error, se_0005$approx_val)) +
  geom_point() +
  geom_hline(yintercept = 1/3, color = "royalblue4") +
  ylab("Approximated Value") + xlab("Standard Error") +
  ylim(c(0.32, 0.34)) + scale_x_reverse()
```

# R's Given Value 

```{r}
integrate(given_function, 0, 1)
```


